---
title: "Statistical Inference Project"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Project Part 1: Simulation Exercise

In this exercise, we begin by simulating results from the exponential distribution.  Each sampling consists of 40 random variables drawn from an exponential distribution with lambda = 0.2, and the samples are taken 1000 times.  The random seed is set (1234) to ensure reproducibility.i

``` {r} 
set.seed(123456)
mean_expos <- NULL
lambda <- 0.2
n <- 40

for(i in 1:1000) {
	mean_expos <- rbind(mean_expos, mean(rexp(n, lambda)))
}
```

We can then plot the histogram of the means.  The mean of the distribution of means should trend toward the mean of the original distribution, which is 1/lamba = 1/0.2 = 5 for this distrubution.  

Theoretical mean:
``` {r}
1/lambda
```

Means of experimental means:
``` {r}
mean(mean_expos)
```

In this histogram plot of the simulated means, the theoretical mean of the distribution is indicated by the red line.

``` {r}
hist(mean_expos, xlab = "Sample mean", main = "Distribution of sample means")
abline(v = 1/lambda, col = "red", lwd = 6)
```

The variance of the sample means is given by
``` {r}
var(mean_expos)
```
While the theoretical variance is given by the variance of the original distribution (var = (1/0.2)^2) divided by the number of observations in each sample set (n = 40)
``` {r}
((1/lambda)^2)/(n)
```

Finally, it can be shown that the resulting distribution of means is approximately normal by plotting the histogram of sample means against a normal distribution with the theoretical mean and standard deviation.
``` {r}
hist(mean_expos, freq = F, xlab = "Sample mean", main = "Distribution of sample means")
x <- seq(min(mean_expos), max(mean_expos), length = 1000)
y <- dnorm(x, mean = 1/lambda, sd = (1/lambda)/sqrt(n))
lines(x,y, col = "blue")
```
The resulting close fit follows from the central limit thereom, which states the distribution of a sufficiently large number of means of any sample will trend towards a normal distribution with mean = mean(original distribution) and sd = sd(original distribution)/(number of samples in each set)

# Project Part 2: Inferential Data Analysis